---
title: "Reproducible Research - Course Project 1"
author: "César Carrera"
date: "6/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning=FALSE)
```

## Introduction

This document containts the scritps require for the Course Project 1 of the Reproducible Research course, acording to the instructions.


## Reading and processing

Loading libraries
```{r, echo=FALSE}
library(ggplot2)
library(tidyverse)
library(lubridate)
library(lattice)
options(scipen=999)
```


Reading the data:

```{r}
data <- read.csv("activity.csv")
str(data)
```
Processing: We need to format the date variable and create the identifier for weekday and weekend.

```{r}
data <- data %>%
    mutate(date_format = ymd(data$date))
```


##Histogram of the total number of steps taken each day
First, we calculate the total number of steps taken per day.

```{r}
data <- data %>%
    group_by(date_format) %>%
    mutate(steps_by_day = sum(steps, na.rm = TRUE)) %>%
    ungroup()

data$steps_by_day[is.na(data$steps)] <- NA_real_
```

The histogram for the steps taken per day is: 
```{r}
hist(data$steps_by_day, main = "Total number of steps taken per day", xlab = "Steps per day", col="green")
```

The mean of the total number of steps taken per day is **`r round(mean(data$steps_by_day),digits=2)`** and the median of the total number of steps taken per day is **`r round(median(data$steps_by_day),digits=2)`**.


##Average number of steps taken

First, we calculate the total number of steps taken per day.

```{r}
average_steps <- data %>%
    group_by(interval) %>%
    summarise(average_by_interval = mean(steps, na.rm = TRUE)) %>%
    ungroup()
```


The time series plot of the average number of steps taken is the following:
```{r}
plot(average_steps$interval, average_steps$average_by_interval, type = "l", lwd = 2, col="green", xlab="Interval", ylab="Average number of steps", main="Average number of steps per intervals")
```

The interval with the maximum number of steps is: **`r average_steps[which.max(average_steps$average_by_interval), ]$interval`**.

##Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
summary(data)
```
There are 2304 cases of NA's in the variable steps. 

The filling in all of the missing values in the dataset, for the variable steps, is goint to be the mean average number of steps by day

```{r}
data2 <- data %>%
    group_by(date_format) %>%
    mutate(steps_by_day = sum(steps, na.rm = TRUE)) %>%
    ungroup()

data2 <- data2 %>%
    mutate(steps_no_na = case_when(!is.na(steps) ~ steps, TRUE ~ steps_by_day))

summary(data2)
```
As shown, the variable steps_no_na has no NA's. Now we plot the histogram for the new variable:
```{r}
data2 <- data2 %>%
    group_by(date_format) %>%
    mutate(steps_by_day_no_na = sum(steps_no_na, na.rm = TRUE)) %>%
    ungroup()

hist(data2$steps_by_day_no_na, main = "Total number of steps taken per day (imputing NA's)", xlab = "Steps per day", col="green")
```

The mean of the total number of steps taken per day (with no NA's) is **`r round(mean(data2$steps_by_day_no_na),digits=2)`** and the median of the total number of steps taken per day is **`r round(median(data2$steps_by_day_no_na),digits=2)`**.


##Differences in activity patterns between weekdays and weekends

First, we create the factor variable for weekday and weekend:
```{r}
data2 <- data2 %>%
    mutate(day_name = weekdays(date_format),
           day_type = case_when(day_name=="sábado" | day_name=="domingo" ~ "Weekend", TRUE ~ "Weekday"))
```

Next we plot the time series of the 5-minute interval and the average number of steps taken, average across all weekday days or weekend day.

```{r}
data2 <- data2 %>%
    group_by(interval, day_type) %>%
    mutate(steps_by_day_type = sum(steps_no_na, na.rm = TRUE)) %>%
    ungroup()

xyplot(steps_by_day_type ~ interval | day_type, data2, type = "l", layout = c(1, 2), 
    xlab = "Interval", ylab = "Average number of steps")
```





